Agroecological Importance of Smooth Brome in Managing Wheat Stem Sawfly (Hymenoptera: Cephidae) via Its Associated Braconid Parasitoids

### 2024-06-13 qmd/projects/msu/wss/paper1.rmd
update

Wheat stem sawfly, *Cephus cinctus* Norton (WSS), causes significant damage in cereal crops throughout the Northern Great Plains of North America. Bracon cephi and Bracon lissogaster are native WSS parasitoids important in managing WSS outbreaks and damage. Smooth Brome (Bromus inermis) grass has been found to be an effective parasitoid sink, and potential trap crop, when grown in areas surrounding wheat fields in Montana. As climate changes continue to alter central and northern Montana, WSS will continue to cause immense damage to dry-land cereal crop yields. My study observes the utilization of smooth brome in providing year-over-year host refuge for Bracon spp. We measured the WSS larval survival rate within controlled smooth brome plots, observing a maximum end-of-year survival of 5.7%. We also collected stems from central and northern Montana where we measured the WSS infestation and parasitoid prevalence within cultivated wheat and adjacent smooth brome. Montana fields sites showed similar high year-end WSS larval mortality while showcasing high levels of parasitism. This research underlines the importance of providing a sustainable ecological buffer for WSS parasitoids as climate conditions continue to change.

@@

Smooth brome (*Bromus inermis* Leyss) is a Eurasian cool-season rhizomatous grass introduced to the United States in 1884 [@Bolwahn2011].*B. inermis* was originally thought of as potentially useful as a trap crop in order to combat the growing WSS problem with cultural control practices [@Criddle1922; @Farstad1945; @Beres2011a]. Due to its low agricultural value, and propensity to form monocultures and out compete many native grasses, *B. inermis* is categorized as invasive in some habitats [@Carlson1985; @Willson2000; @Dillemuth2008]. However, *B. inermis* still holds considerable ecological and potential economic value due to its late senescence and WSS antibiosis properties [@Otfinowski2006; @Rand2024].

@@

All data for experiments was analyzed using R Studio (R Studio version 2024.04.0+735, R 2024).

### Weather data and NDVI analysis

*Weather data.* We assessed the medium- and long-term temperature and precipitation trends of our field sites using weather data from the National Oceanic and Atmospheric Administration (NOAA, Silver Spring, MA, USA). Data for each field site was averaged from three of the closest weather stations to that location. Data was plotted using R package 'ggplot' (version 3.4.4) [@ggplot]. Data was then fit using a linear model using the 'lm' command using average yearly precipitation (inches) as the response variable and year as the predictor.

*NDVI analysis.* We compared the relative greening throughout the growing season between wheat fields and adjacent *B. inermis* using the normalized difference vegetation index (NDVI). NDVI is typically used to assess vegetation health and density, and is calculated from the visible and near-infrared light reflected by vegetation [@Pettorelli2005]. NDVI data measurements were directly downloaded using Google Earth Engine (Google Inc. 2023, Mountain View, CA, USA). Similar to our weather data analysis, NDVI data was modeled using the linear modeling package 'lm' in R to discern differences in NDVI measurments over time.

@@

*Data Collection.* In late August, *B. inermis* stems were collected from each cage. Each stem was sliced open using X-Acto® knives to collect data on infestation, dead larvae and live larvae, as well as very rare parasitism events, for each internode.

@@

*Stem collection and processing.* We conducted a field survey to assess *C. cinctus* infestation, larval mortality, and *B. cephi* and *B. lissogaster* prevalence within *B. inermis* and adjacent wheat fields. Sites were chosen across 2 counties in north-central Montana, United States. (Chouteau, Judith Basin), which consistently experience high *C. cinctus* pressure. Samples were collected from wheat fields and adjacent *B. inermis* in early July and late August in 2022, and 2023 from sites in Big Sandy and Moccasin MT, USA. Sampling sites were set up as 100$m^2$ polygons along the edge of adjoining wheat fields. Four collection squares of 1ft x 1ft were randomly selected within each polygon during both collection events each year. All stems within each 1 x 1 ft square were collected using a shovel to remove both stem and root material. Wheat stems were collected at distances of 5 and 20 meters from the edge of the field. Samples were collected in 4 rows at 10 meters apart. 2 samples were collected in each row at distances of 5 and 20 meters. 1 ft samples were collected at each point.

@@

*Statistical Analysis.* We used generalized linear mixed models with binomial errors (logit link) fit using bound optimization by quadratic approximation, with a maximum of 200,000 iterations, were run in the *lme4* package in R [@lme4] to examine the effects of location and year (fixed effects) on each of the three response variables: proportion of stems infested, proportion of stems cut, and proportion of stems parasitized by *Bracon* spp.

@@

@@

```{r libraries}
library(markdown)
library(tidyverse)
library(glue)
library(lubridate)
library(ggplot2)
library(maps)
library(readxl)
library(kableExtra)
library(openxlsx)
library(lme4)
```

### Historical Weather Analysis & NDVI

@@

We observed a significant linear relationship (*r = `r round(both_prcp_lm$r.squared,3)`, P = `r round(both_prcp_lm$coefficients[2,4],3)`, estimate = `r round(both_prcp_lm$coefficients[2,1],3)`*) between average precipitation and year for both Big Sandy and Moccasin, Montana. For each location and one-year increase, we observed a 0.05 inch decrease in precipitation. ([Fig. 1](#figure1)). Additionally, we observed a significant positive linear relationship (*r = `r round(both_temp_lm$r.squared,3)`, P \< 0.05, estimate = `r round(both_temp_lm$coefficients[2,1],3)`*) between year and average yearly temperature for Moccasin and Big Sandy, MT. Thus, each year, the average daily temperature has increased by `r round(both_temp_lm$coefficients[2,1],2)`°C. ([Fig. 2](#figure2)).

@@

label_x <- as.Date("2023-07-20")
label_y <- 0.75
line_x <- as.numeric(as.Date("2023-07-20"))

bs_ndvi_plot <-
lochy %>%
  ggplot() +
  geom_point(aes(x = date, y = ndvi, color = plant_type),
             alpha = 0.5,
             pch = 1,
             size = 0.75) +
  geom_line(aes(x = date, y = ndvi, color = plant_type, linetype = plant_type),
            alpha = 0.5,
            size = 0.5) +
  geom_smooth(aes(x = date, y = ndvi, color = plant_type),
              se = FALSE) +
  geom_vline(xintercept = line_x,
             linetype = "dashed", color = "black") +
  geom_label(aes(x = label_x,
                 y = 0.75),
             label = "Spring wheat harvest",
             hjust = -0.1,
             size = 3) +
  geom_segment(aes(x = as.Date("2023-07-20")+5, 
                   y = 0.75, 
                   xend = as.Date("2023-07-20"), 
                   yend = 0.75)) +
  theme_bw() +
  guides(color = FALSE) +
  labs(x = "2023 Growing Season",
       y = "NDVI",
       linetype = NULL) +
  scale_linetype_discrete(labels = c(expression(italic("B. inermis")),
                                                      "Spring Wheat"))
```

We observed a notable difference in NDVI when comparing adjacent *B. inermis* and spring wheat. We saw a significant difference in the July NDVI (*`r round(ndvi_model$r.squared,3)`, P \< 0.05*). The *B. inermis* NDVI remained relatively linear in it's downslope (BROME SLOPE POST JULY) compared to the wheat field (WHEAT FIELD POST JULY). ([Fig. 3](#figure3))

Need to add map showing where NDVI measurements came from

Initially, *C. cinctus* heavily infested stems of *B. inermis* in the controlled test conditions. Averaged across both years, we observed 66.5% of stems infested for high treatments and 47.3% of stems for low treatments. We found strong evidence suggesting that there was a significant difference between infestation at high and low treatment levels when holding year constant (*r = `r round(pf_inf_lm$r.squared,3)`, P \< 0.05*, [Fig. 3](#figure4)).

*C. cinctus* infested stems often contained larvae burrowing through multiple nodes. In 2022, `r round(pf_nodes_over_4_table[1,2],3)*100`% of stems were observed to have at least 4 nodes burrowed through, or 5 internodes occupied. In 2023, `r round(pf_nodes_over_4_table[2,2],3)*100`% of stems had at least 4 nodes infested [Fig. 5](#figure5)

```{r stubble data}
# stubble 2023
stubble.2023 <-
  read_excel("/Users/jacksonstrand/Library/CloudStorage/OneDrive-MontanaStateUniversity/Data/BWW/Field/field Sites/2023/2023\ Field Site\ Data.xlsx", 
           sheet = "stubble") %>%
  dplyr::select(-c(name,date.proc,"%inf", "%cut")) %>%
  na.omit() %>%
  mutate(year = "2023")

@@

We quantified *C. cinctus* and *Bracon* spp. parasitoid presence in 6,148 wheat and grass stems across 10 research sites in 2022 and 2023. Initial infestation of *B. inermis* by *C. cinctus* varied between collection sites, with the greatest infestation taking place within our three Big Sandy, MT sampling sites (2023: 65.4%, 2022: 63.1%), while the lowest infestation observed was in Moccasin, MT (2023: 40.8%, 2022: 60.7%). Across all sites in Big Sandy and Moccasin, we observed an average infestation of 57.5% within *B. inermis* and 47.6% within the adjacent wheat. Cutting by *C. cintus* was observed at an average rate of 5.59% (se = 0.02) in *B. inermis* and 47.1% (se = 0.04) in the adjacent cultivated wheat ([Fig. 7](#figure7).

To generate a relative estimate of the number of available *C. cinctus* host larvae within *B. inermis* late in the growing season, we calculated the number of stems that contained at least four or more nodes with burrowing injury. In Big Sandy, 76.2% of collected stems in the fall exhibited at least 4 nodes of burrowing injury at our second sampling date. In Moccasin, the corresponding figure was 66.5%.

@@

```{r infestation_per, eval = F}
comb.brome %>%
  group_by(location, collection, year) %>%
  # filter(collection == "fall") %>%
  reframe(inf = sum(inf)/sum(total.stems)) %>%
  distinct() %>%
  mutate(region = if_else(location %in% c("har", "sun", "k", "chs"), "moccasin", "big sandy")) %>%
  group_by(collection, year, region) %>%
  reframe(inf = mean(inf)) %>%
  na.omit() %>%
  arrange(desc(inf))
```

@@

We utilized the ratio of parasitized stems to cut stems to better understand the true impact of our *B. inermis* sink/source ecology. The largest ratio observed was in Big Sandy, MT in the fall of 2022, where we observed a ratio of 16 parasitoids to a single *C. cinctus*-cut stem. Thus, for every square meter of *B. inermis*, we would observe 16 stems that produced a parasitoid for every 1 stem that was likely to produce an adult *C. cinctus*. All other sample locations also showed positive ratios of parasitoids to cut stems [(Fig. 8)](#figure8).

@@

```{r montana map}
world_map <- map_data("world")
states <- map_data("state") %>%
  filter(region == "montana")
counties <- map_data("county") %>%
  filter(region == "montana")

montana_map <-
ggplot() +
  # geom_polygon(data = world_map, aes(x = long, y = lat, group = group), 
  #              fill = "white", color = "black") +
  geom_polygon(data = counties, aes(x = long, y = lat, group = group),
               fill = NA, color = "black", linewidth = 0.2) +
  geom_polygon(data = states, aes(x = long, y = lat, group = group),
               fill = NA, color = "black", linewidth = 0.9) +
  geom_label(data = coordinates, aes(x = longs, y = lats), 
             color = "black", 
             label = area,
             size = 5,
             vjust = 0,
             hjust = -0.1,
             label.padding = unit(0.25, "lines")) +
  geom_point(data = coordinates, aes(x = longs, y = lats),
             color = "red") +
  coord_cartesian(xlim = c(-115.75, -104.5), ylim = c(44.5, 49)) +
  theme_void() +
  # theme(panel.grid.major = element_blank(),
  #       panel.grid.minor = element_blank()) +
  theme(aspect.ratio = 1/1.8)

## Discussion

>>>>

Wheat stem sawfly, *Cephus cinctus* Norton (WSS), causes significant damage in cereal crops throughout the Northern Great Plains of North America. *Bracon cephi* and *B. lissogaster* are native WSS parasitoids important in managing WSS outbreaks and damage. Smooth Bromegrass (*Bromus inermis*) grass has been found to be an effective parasitoid sink, and potential trap crop, when grown in areas surrounding wheat fields in Montana. As climate changes continue to alter central and northern Montana, WSS will continue to cause immense damage to dry-land cereal crop yields. We analyzed both local temperature and precipitation trends at our sample sites, and found statistically significant increases in temperature and decreases in precipitation over the past fifty years. Our study observes the utilization of smooth brome in providing year-over-year host refuge for Bracon spp parasitoids. We measured the WSS larval infestation and survival rate within controlled smooth brome plots, observing a maximum infestation of 66.5% and a maximum end-of-year survival of 5.7%. We also collected stems from sites in central and northern Montana where we measured the WSS infestation and parasitoid prevalence within cultivated wheat and adjacent smooth brome. Montana fields sites showed high levels of *C. cinctus* infestation within both *B. inermis* and adjacent wheat crop in Big Sandy (64.5% *B. inermis*, 65.7% adjacent wheat) and Moccasin (50.6%, 38.6%). We observed high year-end WSS larval mortality measured as stem cutting of infested stems with *B. inermis* when compared to adjacent wheat in Big Sandy (2.9% *B. inermis*, 46.5% adjacent wheat) and Moccasin (6.4%, 49.4%). Infested stems exhibited parasitism in both Big Sandy (7.11% *B. inermis*, 8.6% adjacent wheat) and Moccasin (6.11%, 2.09%). This research underlines the importance of providing a sustainable ecological buffer for WSS parasitoids as climate conditions continue to change.

@@

Smooth brome (*Bromus inermis* Leyss) is a Eurasian cool-season rhizomatous grass introduced to the United States in 1884 [@Bolwahn2011].*B. inermis* was originally thought of as potentially useful as a trap crop in order to combat the growing WSS problem with cultural control practices [@Criddle1922; @Seamans1928; @Farstad1945; @Beres2011a]. Due to its low agricultural value, and propensity to form monocultures and out compete many native grasses, *B. inermis* is categorized as invasive in some habitats [@Carlson1985; @Willson2000; @Dillemuth2008]. However, *B. inermis* still holds considerable ecological and potential economic value due to its late senescence and WSS antibiosis properties [@Otfinowski2006; @Rand2024].

@@

@@

*Data Collection.* In late August, *B. inermis* stems were collected from each cage. Each stem was dissected using X-Acto® knives to collect data on infestation, dead larvae and live larvae, as well as very rare parasitism events, for each internode.

@@

*Weather data and NDVI* Field sites were chosen across two counties, Choteau and Judith Basin, within the 'Golden Triangle' of Montana, United States [fig. 1](#figure1), areas that consistently experience high *C. cinctus* pressure. We assessed the medium- and long-term temperature and precipitation trends of our field sites using weather data from the National Oceanic and Atmospheric Administration (NOAA, Silver Spring, MA, USA). Data for each field site was averaged from three of the closest weather stations to that location.

To better understand how precipitation patterns vary throughout the growing season, we looked at local daily precipitation totals from weather stations at our sampling sites in Big Sandy and Moccasin, MT. Data was pulled online from historical WeatherUnderground weather station data [@WeatherUnderground]. Precipitation readings were broken into two categories - dates before July 15 and after September 15, as well as between July 15 and September 15. These dates were chosen based on typical wheat harvest and historical periods of known *Bracon* spp activity. In the 'Golden Triangle', typical spring wheat harvest takes place in mid-July on average [@Wilson1923; @Zhu2021].

We used the normalized difference vegetation index (NDVI) to compare the relative greening throughout the growing season between wheat fields and adjacent *B. inermis* ([fig. 2](#figure2)). NDVI is typically used to assess vegetation health and density, and is calculated from the visible and near-infrared light reflected by vegetation [@Pettorelli2005]. NDVI data measurements were directly downloaded using Google Earth Engine (Google Inc. 2023, Mountain View, CA, USA).

*Stem collection and processing.* We conducted a field survey to assess *C. cinctus* infestation, larval mortality, and *B. cephi* and *B. lissogaster* prevalence within *B. inermis* and adjacent wheat fields. Samples were collected from wheat fields and adjacent *B. inermis* in early July and late August in 2022, and 2023 from sites in Big Sandy and Moccasin MT, USA. Sampling sites were set up as 100$m^2$ polygons along the edge of adjoining wheat fields. Four collection squares of 1ft x 1ft were randomly selected within each polygon during both collection events each year. All stems within each 1 x 1 ft square were collected using a shovel to remove both stem and root material. Wheat stems were collected at distances of 5 and 20 meters from the edge of the field. Samples were collected in 4 rows at 10 meters apart. 2 samples were collected in each row at distances of 5 and 20 meters. 1 ft samples were collected at each point.

@@

@@

### Statistical Analysis

All statistical analyses were done in R [@R2024] using R Studio (version 4.3.2). Data was plotted using R package 'ggplot' (version 3.4.4) [@ggplot]. We analyzed long- and short-term weather trends using linear modeling and the 'lm' command. Weather data was analyzed using either average yearly precipitation (inches) or average yearly temperature (°C) as the response variable and year as the predictor. Similar to our weather data analysis, NDVI data was linear modeled to discern differences in NDVI measurements over time.

We used generalized linear mixed models with binomial errors (logit link) fit using bound optimization by quadratic approximation, with a maximum of 200,000 iterations, were run in the *lme4* package in R [@lme4] to examine the effects of location and year (fixed effects) on each of the three response variables: proportion of stems infested, proportion of stems cut, and proportion of stems parasitized by *Bracon* spp.

@@

```{r libraries}
library(markdown)
library(tidyverse)
library(glue)
library(lubridate)
library(ggplot2)
library(maps)
library(readxl)
library(kableExtra)
library(openxlsx)
library(lme4)
library(gridExtra)
```

*C. cinctus* heavily infested stems of *B. inermis* in the controlled test conditions. Averaged across both years, we observed 66.5% of stems infested for high treatments and 47.3% of stems for low treatments. We found strong evidence suggesting that there was a significant difference between infestation at high and low treatment levels when holding year constant (*r = `r round(pf_inf_lm$r.squared,3)`, P \< 0.05*, [Fig. 3](#figure3)).

*C. cinctus* infested stems often contained larvae burrowing through multiple nodes. In 2022, `r round(pf_nodes_over_4_table[1,2],3)*100`% of stems were observed to have at least 4 nodes burrowed through, or 5 internodes occupied. In 2023, `r round(pf_nodes_over_4_table[2,2],3)*100`% of stems had at least 4 nodes infested [Fig. 4](#figure4)

#### Historical Weather Analysis & NDVI

@@

```{r bs_seasonal_plot}
lochy_precip <-
read_excel("weather/precip_scrape.xlsx", sheet = "big_sandy")

line_table <-
lochy_precip %>%
  dplyr::select(date, prcp) %>%
  mutate(month = month(date),
         day = day(date),
         year = as.factor(year(date))) %>%
  mutate(new_date = make_date(year = 0,
                              month = month,
                              day = day)) %>%
   mutate(category = case_when(new_date < "0000-07-15" ~ "pre",
                              new_date > "0000-09-15" ~ "post",
                              new_date >= "0000-07-15" & new_date <= "0000-09-15" ~ "mid")) %>%
  mutate(model_category = if_else(category == "mid", 
                                  "mid", "no"))

line_table_average <-
  line_table %>%
  group_by(month, day) %>%
  reframe(category, new_date,
          prcp = mean(prcp))

category_averages <-
line_table_average %>%
  na.omit() %>%
  group_by(category) %>%
  reframe(prcp = mean(prcp))

bs_seasonal_lm <-  
line_table %>%
  lm(prcp ~ model_category + year, data = .) %>%
  summary()

mult = 3

smooth_plot <-
ggplot() +
  geom_smooth(aes(x = new_date, y = prcp),
              data = line_table_average,
              se = FALSE,
              method = 'loess',
              color = "black") +
  geom_rect(aes(xmin = as.Date("0000-05-01", 
                               format = "%Y-%m-%d"), 
                xmax = as.Date("0000-07-15", 
                               format = "%Y-%m-%d"),
                ymin = 0, 
                ymax = 0.0528 * mult),
            alpha = 0.1,
            fill = "blue") +
  geom_segment(aes(x = as.Date("0000-05-01", 
                               format = "%Y-%m-%d"), 
                   xend = as.Date("0000-07-15", 
                                  format = "%Y-%m-%d"),
                   y = 0.0528 * mult, yend = 0.0528 * mult),
               color = "blue",
               alpha = 0.5) +
  geom_rect(aes(xmin = as.Date("0000-07-15", 
                               format = "%Y-%m-%d"), 
                xmax = as.Date("0000-09-15", 
                               format = "%Y-%m-%d"),
                ymin = 0, 
                ymax = 0.0164 * mult),
            alpha = 0.1,
            fill = "blue") +
  geom_segment(aes(x = as.Date("0000-07-15", 
                               format = "%Y-%m-%d"), 
                   xend = as.Date("0000-09-15", 
                                  format = "%Y-%m-%d"),
                   y = 0.0164 * mult, yend = 0.0164 * mult),
               color = "blue",
               alpha = 0.5) +
  geom_rect(aes(xmin = as.Date("0000-09-15", 
                               format = "%Y-%m-%d"), 
                xmax = as.Date("0000-10-31", 
                               format = "%Y-%m-%d"),
                ymin = 0, 
                ymax = 0.0281 * mult),
            alpha = 0.1,
            fill = "blue") +
  geom_segment(aes(x = as.Date("0000-09-15", 
                               format = "%Y-%m-%d"), 
                   xend = as.Date("0000-10-31", 
                                  format = "%Y-%m-%d"),
                   y = 0.0281 * mult, yend = 0.0281 * mult),
               color = "blue",
               alpha = 0.5) +
  theme_bw()

bs_seasonal_plot <-
ggplot() +
  geom_line(aes(x = new_date, y = prcp),
            data = line_table_average %>%
              filter(category == "pre"),
            color = "black",
            linewidth = 0.25) +
  geom_line(aes(x = new_date, y = prcp),
            data = line_table_average %>%
              filter(category == "post"),
            color = "black",
            linewidth = 0.25) +
  geom_line(aes(x = new_date, y = prcp),
            data = line_table_average %>%
              filter(category == "mid"),
            color = "black") +
  geom_rect(aes(xmin = as.Date("0000-05-01", 
                               format = "%Y-%m-%d"), 
                xmax = as.Date("0000-07-15", 
                               format = "%Y-%m-%d"),
                ymin = 0, 
                ymax = 0.0528 * mult),
            alpha = 0.1,
            fill = "blue") +
  geom_segment(aes(x = as.Date("0000-05-01", 
                               format = "%Y-%m-%d"), 
                   xend = as.Date("0000-07-15", 
                                  format = "%Y-%m-%d"),
                   y = 0.0528 * mult, yend = 0.0528 * mult),
               color = "blue",
               alpha = 0.5) +
  geom_rect(aes(xmin = as.Date("0000-07-15", 
                               format = "%Y-%m-%d"), 
                xmax = as.Date("0000-09-15", 
                               format = "%Y-%m-%d"),
                ymin = 0, 
                ymax = 0.0164 * mult),
            alpha = 0.1,
            fill = "blue") +
  geom_segment(aes(x = as.Date("0000-07-15", 
                               format = "%Y-%m-%d"), 
                   xend = as.Date("0000-09-15", 
                                  format = "%Y-%m-%d"),
                   y = 0.0164 * mult, yend = 0.0164 * mult),
               color = "blue",
               alpha = 0.5) +
  geom_rect(aes(xmin = as.Date("0000-09-15", 
                               format = "%Y-%m-%d"), 
                xmax = as.Date("0000-10-31", 
                               format = "%Y-%m-%d"),
                ymin = 0, 
                ymax = 0.0281 * mult),
            alpha = 0.1,
            fill = "blue") +
  geom_segment(aes(x = as.Date("0000-09-15", 
                               format = "%Y-%m-%d"), 
                   xend = as.Date("0000-10-31", 
                                  format = "%Y-%m-%d"),
                   y = 0.0281 * mult, yend = 0.0281 * mult),
               color = "blue",
               alpha = 0.5) +
  labs(x = NULL,
       y = "Average Daily Precipitation (in)",
       color = NULL) +
  scale_y_continuous(sec.axis = sec_axis(~ . * mult, 
                                         name = "Average Seasonal Precpitation (in)")) +
  scale_x_date(
    date_breaks = "1 month",
    date_labels = "%B") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```{r moc_seasonal_plot}
moc_precip <-
read_excel("weather/precip_scrape.xlsx", sheet = "moccasin")

line_table <-
moc_precip %>%
  dplyr::select(date, prcp) %>%
  mutate(month = month(date),
         day = day(date),
         year = as.factor(year(date))) %>%
  mutate(new_date = make_date(year = 0,
                              month = month,
                              day = day)) %>%
   mutate(category = case_when(new_date < "0000-07-15" ~ "pre",
                              new_date > "0000-09-15" ~ "post",
                              new_date >= "0000-07-15" & new_date <= "0000-09-15" ~ "mid")) %>%
  mutate(model_category = if_else(category == "mid", 
                                  "mid", "no"))

line_table_average <-
  line_table %>%
  group_by(month, day) %>%
  reframe(category, new_date,
          prcp = mean(prcp))

category_averages <-
line_table_average %>%
  na.omit() %>%
  group_by(category) %>%
  reframe(prcp = mean(prcp))

moc_seasonal_lm <-  
line_table %>%
  lm(prcp ~ model_category + year, data = .) %>%
  summary()

mult = 3

moc_seasonal_plot <-
ggplot() +
  geom_line(aes(x = new_date, y = prcp),
            data = line_table_average %>%
              filter(category == "pre"),
            color = "black",
            linewidth = 0.25) +
  geom_line(aes(x = new_date, y = prcp),
            data = line_table_average %>%
              filter(category == "post"),
            color = "black",
            linewidth = 0.25) +
  geom_line(aes(x = new_date, y = prcp),
            data = line_table_average %>%
              filter(category == "mid"),
            color = "black") +
  geom_rect(aes(xmin = as.Date("0000-05-01", 
                               format = "%Y-%m-%d"), 
                xmax = as.Date("0000-07-15", 
                               format = "%Y-%m-%d"),
                ymin = 0, 
                ymax = 0.0528 * mult),
            alpha = 0.1,
            fill = "blue") +
  geom_segment(aes(x = as.Date("0000-05-01", 
                               format = "%Y-%m-%d"), 
                   xend = as.Date("0000-07-15", 
                                  format = "%Y-%m-%d"),
                   y = 0.0528 * mult, yend = 0.0528 * mult),
               color = "blue",
               alpha = 0.5) +
  geom_rect(aes(xmin = as.Date("0000-07-15", 
                               format = "%Y-%m-%d"), 
                xmax = as.Date("0000-09-15", 
                               format = "%Y-%m-%d"),
                ymin = 0, 
                ymax = 0.0164 * mult),
            alpha = 0.1,
            fill = "blue") +
  geom_segment(aes(x = as.Date("0000-07-15", 
                               format = "%Y-%m-%d"), 
                   xend = as.Date("0000-09-15", 
                                  format = "%Y-%m-%d"),
                   y = 0.0164 * mult, yend = 0.0164 * mult),
               color = "blue",
               alpha = 0.5) +
  geom_rect(aes(xmin = as.Date("0000-09-15", 
                               format = "%Y-%m-%d"), 
                xmax = as.Date("0000-10-31", 
                               format = "%Y-%m-%d"),
                ymin = 0, 
                ymax = 0.0281 * mult),
            alpha = 0.1,
            fill = "blue") +
  geom_segment(aes(x = as.Date("0000-09-15", 
                               format = "%Y-%m-%d"), 
                   xend = as.Date("0000-10-31", 
                                  format = "%Y-%m-%d"),
                   y = 0.0281 * mult, yend = 0.0281 * mult),
               color = "blue",
               alpha = 0.5) +
  labs(x = NULL,
       y = "Average Daily Precipitation (in)",
       color = NULL) +
  scale_y_continuous(sec.axis = sec_axis(~ . * mult, 
                                         name = "Average Seasonal Precpitation (in)")) +
  scale_x_date(
    date_breaks = "1 month",
    date_labels = "%B") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

In both Big Sandy and Moccasin, MT, we observed a decrease in average yearly precipitation and increase in average yearly temperature. We observed a significant linear relationship (*r = `r round(both_prcp_lm$r.squared,3)`, P = `r round(both_prcp_lm$coefficients[2,4],3)`, estimate = `r round(both_prcp_lm$coefficients[2,1],3)`*) between average precipitation and year for both Big Sandy and Moccasin, Montana. For each location and one-year increase, we observed a 0.05 inch decrease in precipitation. ([Fig. 5](#figure5)). Additionally, we observed a significant positive linear relationship (*r = `r round(both_temp_lm$r.squared,3)`, P \< 0.05, estimate = `r round(both_temp_lm$coefficients[2,1],3)`*) between year and average yearly temperature for Moccasin and Big Sandy, MT. Thus, each year, the average daily temperature has increased by `r round(both_temp_lm$coefficients[2,1],2)`°C. ([Fig. 6](#figure6)). 
Analyzing local, precipitation data over the past five years yielded information on how central and northern Montana experience relative dryness during the late-summer months. In Big Sandy, when controlling for year, we saw strong evidence of a significant difference (*p<0.05, F=2.067*, [fig. 7](#figure7)) between daily rainfall in months prior July 15 and after September 15 compared to days within the July 15 to September 15 'late-summer' period. Analysis of Moccain precipitation data revealed that there is little evidence (*p=0.08, F=1.535*) to suggest a difference between the two date categories.

@@

label_x <- as.Date("2023-07-15")
label_y <- 0.75
line_x <- as.numeric(as.Date("2023-07-15"))

bs_ndvi_plot <-
lochy %>%
  ggplot() +
  geom_point(aes(x = date, y = ndvi, color = plant_type),
             alpha = 0.5,
             pch = 1,
             size = 0.75) +
  geom_line(aes(x = date, y = ndvi, 
                color = plant_type, linetype = plant_type),
            alpha = 0.5,
            size = 0.5) +
  geom_smooth(aes(x = date, y = ndvi, color = plant_type),
              se = FALSE) +
  geom_vline(xintercept = line_x,
             linetype = "dashed", color = "black") +
  geom_label(aes(x = label_x,
                 y = 0.8),
             label = "Average spring wheat harvest",
             hjust = -0.1,
             size = 3) +
  theme_bw() +
  guides(color = FALSE) +
  labs(x = NULL,
       y = "NDVI",
       linetype = NULL) +
  scale_linetype_discrete(labels = c(expression(italic("B. inermis")),"Spring Wheat"))
```

We observed a notable difference in NDVI when comparing adjacent *B. inermis* and spring wheat. We saw a significant difference in the *B. inermis* post-July 15 NDVI (*`r round(ndvi_model$r.squared,3)`, P<0.05*) when compared to the post-July 15 NDVI of the adjacent wheat crop. The *B. inermis* NDVI remained relatively linear in it's downslope compared to the wheat field ([Fig. 7](#figure7)).

#### Field Site Samples
```{r stubble data}
# stubble 2023
stubble.2023 <-
  read_excel("/Users/jacksonstrand/Library/CloudStorage/OneDrive-MontanaStateUniversity/Data/BWW/Field/field Sites/2023/2023\ Field Site\ Data.xlsx", 
           sheet = "stubble") %>%
  dplyr::select(-c(name,date.proc,"%inf", "%cut")) %>%
  na.omit() %>%
  mutate(year = "2023")

@@

We quantified *C. cinctus* and *Bracon* spp. parasitoid presence in 6,148 wheat and grass stems across 10 research sites in 2022 and 2023. Initial infestation of *B. inermis* by *C. cinctus* varied between collection sites, with the greatest infestation taking place within our three Big Sandy, MT sampling sites (2023: 65.4%, 2022: 63.1%), while the lowest infestation observed was in Moccasin, MT (2023: 40.8%, 2022: 60.7%). Across all sites in Big Sandy and Moccasin, we observed an average infestation of 57.5% within *B. inermis* and 47.6% within the adjacent wheat. Cutting by *C. cintus* was observed at an average rate of 5.59% in *B. inermis* and 47.1% in the adjacent cultivated wheat ([Fig. 8](#figure8).

To generate a relative estimate of the number of available *C. cinctus* host larvae within *B. inermis* late in the growing season, we calculated the number of stems that contained at least four or more nodes with burrowing injury. In Big Sandy, 76.2% of collected stems in the fall exhibited at least 4 nodes of burrowing injury at our second sampling date. In Moccasin, the corresponding figure was 66.5% ([Fig. 9](#figure9)).

@@

```{r per.par}
comb.stubble %>%
  mutate(region = if_else(location %in% c("har", "sun", "k", "chs"), "moccasin", "big sandy")) %>%
  group_by(region) %>%
  reframe(total.stems = sum(total.stems),
          inf = sum(inf),
          cut = sum(cut),
          par = sum(c, e),
          inf.per = inf/total.stems,
          cut.per = cut/inf,
          par.per = par/inf) %>%
  dplyr::select(par.per)

comb.stubble %>%
  dplyr::select(-c(uninf, cut, c, e, total.stems))
```

```{r infestation_per, eval = F}
comb.brome %>%
  group_by(location, collection, year) %>%
  # filter(collection == "fall") %>%
  reframe(inf = sum(inf)/sum(total.stems)) %>%
  distinct() %>%
  mutate(region = if_else(location %in% c("har", "sun", "k", "chs"), "moccasin", "big sandy")) %>%
  group_by(collection, year, region) %>%
  reframe(inf = mean(inf)) %>%
  na.omit() %>%
  arrange(desc(inf))

plot.data %>%
  filter(collection == "fall") %>%
  group_by(region, collection, plant_type) %>%
  reframe(per.par = sum(par)/sum(total.stems))

comb.brome %>%
  mutate(region = if_else(location %in% c("har", "sun", "k", "chs"), "moccasin", "big sandy")) %>%
  na.omit() %>%
  group_by(region, collection) %>%
  reframe(inf = sum(inf),
          cut = sum(cut),
          cut.per = cut/inf)
```

@@

We utilized the ratio of parasitized stems to cut stems to better understand the true impact of our *B. inermis* sink/source ecology. The largest ratio observed was in Big Sandy, MT in the fall of 2022, where we observed a ratio of 16 parasitoids to a single *C. cinctus*-cut stem. Thus, for every square meter of *B. inermis*, we would observe 16 stems that produced a parasitoid for every 1 stem that was likely to produce an adult *C. cinctus*. All other sample locations also showed positive ratios of parasitoids to cut stems [(Fig. 10)](#figure10).

@@

## Discussion
<!-- CONTROLLED INFESTATION -->
Assessment of *C. cicntus* infestation and larval mortality within *B. inermis* from out controlled infestation survey showed high infestation coupled with high larval mortality prior to overwintering diapause. Even in cases of high infestation (70%+), larval survival to cutting topped out around around 15%, pointing towards *B. inermis'* ability to serve as a quality sink for *C. cinctus* larvae. These results confirm the assertions by Criddle (1992) and Seamans (1928) that *B. inermis* can serve as a quality *C. cinctus* trap crop. In addition to high larval mortality, stem dissection exhibited high numbers of stems with over four internodes burrowed through by *C. cinctus* larvae. Larval movement through stems can be quick in the case of susceptible cultivated cereal grains [@Criddle1923]. However, *B. inermis*, given it's presumed antibiosis properties, leads to slower larval development and thus movement through the stem. We found several *B.inermis* stems with living larvae as late as late-September.

<!-- WEATHER DATA -->
As the climate of central and northern Montana continues to change, wheat stem sawfly infestation will continue to impact cereal crop yields. Precipitation and temperature data from our sampling sites underscores the importance of continuing to monitor and conserve areas of semi-natural grassland. 50-year trends analyzed in this study show similar results to Pederson et al (2009) in which the dry-land agricultural environments found in central and northern Montana are likely to continue to become hotter and dryer on average. Over the past fifty years, precipitation in Big Sandy has decreased by an average of 0.05 inches per year. Given the trend continues, by 2050 the area could experience an inch less of rain per year.
